---
output: github_document
bibliography: vignettes/rEMM.bib
---

```{r echo=FALSE, results = 'asis'}
pkg <- 'rEMM'

source("https://raw.githubusercontent.com/mhahsler/pkg_helpers/main/pkg_helpers.R")
pkg_title(pkg)
```


Implements TRACDS (Temporal Relationships
between Clusters for Data Streams), a generalization of the
Extensible Markov Model (EMM),
to model transition probabilities in sequence data [@Hahsler+Dunham:2011]. TRACDS adds a temporal or order model
to data stream clustering by superimposing a dynamically adapting
Markov chain. It also provides an implementation of EMM (TRACDS on top of tNN
data stream clustering) [@Hahsler+Dunham:2010]. 

The package also provides interface classes DSC_tNN and DSC_EMM for data stream
mining with [`stream`](https://michael.hahsler.net/stream/) [@Hahsler:2017].


```{r echo=FALSE, results = 'asis'}
pkg_install(pkg)
```

## Usage

We use an artificial dataset with four clusters. Points are generated using the fixed sequence
<1,2,1,3,4> through the four clusters. The lines below indicate the sequence.

```{r example_data}
library(rEMM)

data("EMMsim")

plot(EMMsim_train, pch = NA)
lines(EMMsim_train, col = "gray")
points(EMMsim_train, pch = EMMsim_sequence_train)
```

EMM recovers the components and sequence information. We use EMM and then
recluster the learned structure, assuming that we know there are four
components. The graph below represents a Markov model of the sequence.

```{r example_model}
emm <- EMM(threshold = 0.1, measure = "euclidean")
build(emm, EMMsim_train)
emmc <- recluster_hclust(emm, k = 4, method = "average")
plot(emmc, method = "MDS")
```

We can now score new sequences. Here, we use a test sequence created in the
same way as the training data and calculate the product of the transition
probabilities in the model. A high score indicates a good match.

```{r}
score(emmc, EMMsim_test)
```

## Vignettes

* [Getting started with rEMM](https://michael.hahsler.net/rEMM/articles/rEMM.html)
  introduces building, inspecting, and using an EMM.
* [Adding temporal structure modeling to standard clustering](https://michael.hahsler.net/rEMM/articles/TRAC.html)
  shows how to add a temporal model to clustering methods such as k-means.
* [Adding temporal structure modeling to stream clustering](https://michael.hahsler.net/rEMM/articles/stream.html)
  connects stream clusterers to TRACDS.

# Acknowledgements
    
Development of this package was supported in part by NSF IIS-0948893 and R21HG005912 from
the National Human Genome Research Institute.

# References

::: {#refs}
:::


